---
title: "BSFG"
author: "Sarah Odell"
date: "3/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Running BSFG on Arabidopsis Root Data (data from Wolfgang Busch)
For information on how the data was transformed, see RootDataTransformed.Rmd
This should be run on farm. It takes about 9 hours to build the model and run MCMC using 50 factors and 70 iterations 
```{r libraries, echo=FALSE, eval=FALSE}
library("BSFG")
library("reshape2")
library("ggplot2")
```


Read in data
```{r readdata, eval=FALSE}
roots=read.table('~/Documents/PBGG/Rotations/Runcie/Archive/T_RootTraitMatrix.txt',sep='\t',header=TRUE)
key=read.table('~/Documents/PBGG/Rotations/Runcie/Archive/T_RootTraitPlantKey.txt',sep='\t',header=TRUE)

Y = roots[,-1]
data = key
dim(Y)
```


Set run parameters. Number of factors is initialized at 50
```{r param, eval=FALSE}
run_parameters = BSFG_control(
  scale_Y = TRUE,
  simulation = FALSE,
  h2_divisions = 20,
  h2_step_size = NULL,
  burn = 00,
  k_init = 50
)
```

Set the prior hyperparameters of the BSFG model
```{r prior, eval=FALSE}
priors = BSFG_priors(
  tot_Y_var = list(V = 0.5,   nu = 5),
  tot_F_var = list(V = 18/20, nu = 20),
  Lambda_prior=list(
    sampler=sample_Lambda_prec_ARD,
    Lambda_df=3,
    delta_1   = list(shape = 2,  rate = 1),
    delta_2   = list(shape = 3, rate = 1)),
  B_prior=list(
    sampler=sample_B_prec_ARD,
    global=list(V=1,nu=3),
    global_F=list(V=1,nu=3),
    B_df=3,
    B_F_df=2),
  h2_priors_resids_fun = function(h2s,n) 1,
  h2_priors_factors_fun = function(h2s,n) 1
)
```

##Construct the model
This takes about 20 minutes
```{r model, eval=FALSE}
BSFG_state = BSFG_init(Y,
                       model=~1+(1|ACC_ID),
                       data = data,
                       run_parameters=run_parameters,
                       priors=priors
)
```

I would recommend saving the model to a file so that it can be easily loaded when examining
posteriors later
```{r savemodel, eval=FALSE}
saveRDS(BSFG_state,file="BSFG_state.rds")
```


##MCMC run
```{r mcmc, eval=FALSE}
n_samples = 100;  # how many samples to collect at once?
BSFG_state$run_parameters$simulation=F
for(i  in 1:70) {
  print(sprintf('Run %d',i))
  BSFG_state = sample_BSFG(BSFG_state,n_samples,grainSize=1)  
  # run MCMC chain n_samples iterations. grainSize is a parameter for
  #parallelization (smaller = more parallelization)
  # set of commands to run during burn-in period to help chain converge
  if(BSFG_state$current_state$nrun < BSFG_state$run_parameters$burn || i < 50) {
    BSFG_state = reorder_factors(BSFG_state) # Factor order doesn't "mix" well in the MCMC. We can help it by manually re-ordering from biggest to smallest
    BSFG_state = clear_Posterior(BSFG_state)
    BSFG_state$run_parameters$burn = max(BSFG_state$run_parameters$burn,BSFG_state$current_state$nrun+100) # if you made changes, set a new burn-in period
    print(BSFG_state$run_parameters$burn)
  }
  BSFG_state = save_posterior_chunk(BSFG_state)  # save any accumulated posterior samples in the database to release memory
  print(BSFG_state) # print status of current chain
  plot(BSFG_state) # make some diagnostic plots. These are saved in a pdf booklet: diagnostic_plots.pdf
}
```



##Analyzing the Posterior
Reload the posterior samples that you need. (Reloading the full posterior requires more memory
than we can use on farm)

If you are starting a new session after running MCMC above, you can load everything 
back starting in the director above the BSFG_run directory created by BSFG

```{r posterior, eval=FALSE}
BSFG_state=readRDS("BSFG_state.rds")
BSFG_state$current_state=readRDS('BSFG_run/current_state.rds')
BSFG_state$Posterior=readRDS('BSFG_run/Posterior/Posterior_base.rds')

#This is too big 
#BSFG_state$Posterior = reload_Posterior(BSFG_state)
#Just load the things you need for now
BSFG_state$Posterior$Lambda = load_posterior_param(BSFG_state,'Lambda')
dim(Lambda)
```


```{r otherparams, eval=FALSE}
BSFG_state$Posterior$F_h2=load_posterior_param(BSFG_state,'F_h2')
BSFG_state$Posterior$resid_h2=load_posterior_param(BSFG_state,'resid_h2')
BSFG_state$Posterior$tot_Eta_prec=load_posterior_param(BSFG_state,'tot_Eta_prec')
BSFG_state$Posterior$B=load_posterior_param(BSFG_state,'B')
BSFG_state$Posterior$B_F=load_posterior_param(BSFG_state,'B_F')
BSFG_state$Posterior$U_F=load_posterior_param(BSFG_state,'U_F')

Lambda_mean = get_posterior_mean(BSFG_state,Lambda)
U_F_mean = get_posterior_mean(BSFG_state,U_F)
```

Then we can create plots to look at the posterior 
```{r plots, eval=FALSE}
# boxplots are good ways to visualize Posterior distributions on sets of related parameters
boxplot(BSFG_state$Posterior$F_h2[,1,])

# get posterior distribution on a function of parameters
# This is how to calculate the G-matrix for random effect #1 (ie animal above.)
G_samples = get_posterior_FUN(BSFG_state,Lambda %*% diag(F_h2[1,]) %*% t(Lambda) + resid_h2[1,]/tot_Eta_prec[1,])

# get posterior mean of a parameter
G = get_posterior_mean(G_samples)

# get Highest Posterior Density intervals for paramters
F_h2_HPD = get_posterior_HPDinterval(BSFG_state,F_h2)

boxplot(BSFG_state$Posterior$B[,2,],outline=F);abline(h=0)
boxplot(BSFG_state$Posterior$B_F[,2,],outline=F);abline(h=0)

BSFG_state$Posterior$U_F = load_posterior_param(BSFG_state,'U_F')
U_F = get_posterior_mean(BSFG_state,U_F)

```

We can also start to look at modules.
The following code is from Caryn Johansen ()

```{r g1, eval=FALSE}
#load Posterior lambda array
Lambda=BSFG_state$Posterior$Lambda

#examine the factor loadings across all the Posterior
g1 = Lambda[,1,] #for one trait
g1_m = melt(g1)
colnames(g1_m)=c("posterior","factor","loading")

summary(g1_m)
```

Visualizing the first Posterior
```{r f, eval=FALSE}
f = Lambda[1,,]
f=melt(f)
colnames(f)=c("trait","factor","loading")
f$factor=as.factor(f$factor)

ggplot(f,aes(x=factor,y=loading)) + geom_boxplot()
```

Visualizing the last Posterior
```{r last,eval=FALSE}
n = dim(Lambda)[1]
p=Lambda[n,,]
p=melt(p)
colnames(p)=c("trait","factor","loading")
p$factor=as.factor(p$factor)

ggplot(p,aes(x=factor,y=loading)) + geom_boxplot()
```

Looking at the modules for the last posterior.
The cutoff for loading values is 0.2
```{r modules,eval=FALSE}
modules=p[p$loading >= 0.2]

table(modules$factor)
saveRDS(modules,file="modules.rds")
```


