---
title: "Transform Root Data"
author: "Sarah Odell"
date: "3/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Transformation of Arabidopsis Root data obtained from Wolfgang Busch. Arabidopsis seedlings were grown under
various nutrient programs (pH, S, P, and Fe) and a control. Traits were collected over a 5-day time course.

In order to run BSFG on these traits, they need to be combined into one matrix, with each row being an individual plant. A separate data table is created associating unique plant IDs with their Accession number and nutrient treatment. Relative growth rate traits which appear non-normal are log transformed.

Start in the Archive directory and read in the data files.
```{r libraries, eval=FALSE}
setwd('~/Documents/PBGG/Rotations/Runcie/Archive/')

library('tidyverse')

usms_traits = read.csv('20180219_USMS_traits.csv',header=TRUE)
s_traits = read.csv('20180219_S_traits.csv',header=TRUE)
fe_traits= read.csv('20180219_Fe_traits.csv', header=TRUE)
p_traits=read.csv('20180219_P_traits.csv',header=TRUE)
ph_traits=read.csv('20180219_pH_traits.csv',header=TRUE)
```

The Fe data has a lot of empty rows, so remove these.
```{r empty, eval=FALSE}
fe_traits=fe_traits[rowSums(is.na(fe_traits)) != ncol(fe_traits),]
```

Make unique plant ID for each treatment

```{r plantid, eval=FALSE}
#Fe treatment
fe_traits=arrange(fe_traits,ACC_ID)
fe_by_acc = fe_traits %>%
  group_by(ACC_ID) %>%
  summarise(rep.num=length(ACC_ID))


fe_pid = c()
for(i in 1:dim(fe_by_acc)[1]){
  
  for(j in 1:fe_by_acc[i,]$rep.num){
    fe_pid=rbind(fe_pid,paste0('Fe_',fe_by_acc[i,]$ACC_ID,'_',j))
  }
}
fe_traits$plant_id = fe_pid


#USMS
usms_traits=arrange(usms_traits,ACC_ID)
usms_by_acc = usms_traits %>%
  group_by(ACC_ID) %>%
  summarise(rep.num=length(ACC_ID))


usms_pid = c()
for(i in 1:dim(usms_by_acc)[1]){
  
  for(j in 1:usms_by_acc[i,]$rep.num){
    usms_pid=rbind(usms_pid,paste0('USMS_',usms_by_acc[i,]$ACC_ID,'_',j))
  }
}
usms_traits$plant_id = usms_pid

#pH
ph_traits=arrange(ph_traits,ACC_ID)
ph_by_acc = ph_traits %>%
  group_by(ACC_ID) %>%
  summarise(rep.num=length(ACC_ID))

ph_pid = c()
for(i in 1:dim(ph_by_acc)[1]){
  
  for(j in 1:ph_by_acc[i,]$rep.num){
    ph_pid=rbind(ph_pid,paste0('pH_',ph_by_acc[i,]$ACC_ID,'_',j))
  }
}
ph_traits$plant_id = ph_pid

#S
s_traits=arrange(s_traits,ACC_ID)
s_by_acc = s_traits %>%
  group_by(ACC_ID) %>%
  summarise(rep.num=length(ACC_ID))

s_pid = c()
for(i in 1:dim(s_by_acc)[1]){
  for(j in 1:s_by_acc[i,]$rep.num){
    s_pid=rbind(s_pid,paste0('S_',s_by_acc[i,]$ACC_ID,'_',j))
  }
}
s_traits$plant_id = s_pid

#P
p_traits=arrange(p_traits,ACC_ID)
p_by_acc = p_traits %>%
  group_by(ACC_ID) %>%
  summarise(rep.num=length(ACC_ID))

p_pid = c()
for(i in 1:dim(p_by_acc)[1]){
  for(j in 1:p_by_acc[i,]$rep.num){
    p_pid=rbind(p_pid,paste0('P_',p_by_acc[i,]$ACC_ID,'_',j))
  }
}
p_traits$plant_id = p_pid
```

Making transformed rates datatable. log_traits is a vector of all the traits that appeared strongly non-normal and that were effectively normalized by log-transformation.
```{r log, eval=FALSE}
log_traits=c("USMS_Root_tortuosity_day001","USMS_Root_tortuosity_day002","USMS_Root_tortuosity_day003",
             "USMS_Root_tortuosity_day004","USMS_Root_tortuosity_day005","USMS_Root_tortuosity_avg",
             "USMS_Relative_root_growth_rate_day001_day002","USMS_Relative_root_growth_rate_day002_day003",
             "USMS_Relative_root_growth_rate_day003_day004","USMS_Relative_root_growth_rate_day004_day005",
             "USMS_Relative_root_growth_rate_avg","P_GR_TL_day_1.2","P_GR_TL_day_2.3","P_GR_TL_day_3.4",
             "P_GR_TL_day_4.5","P_GR_TL_AVG","P_sdX_sdY_per_Pix_day_1","P_sdX_sdY_per_Pix_day_2","P_sdX_sdY_per_Pix_day_3",
             "P_sdX_sdY_per_Pix_day_4","P_sdX_sdY_per_Pix_day_5","P_sdX_sdY_per_Pix_AVG", "S_GR_TL_day_1.2",
             "S_GR_TL_day_2.3","S_GR_TL_day_3.4","S_GR_TL_day_4.5","S_GR_TL_AVG","S_sdX_sdY_per_Pix_day_1",
             "S_sdX_sdY_per_Pix_day_2","S_sdX_sdY_per_Pix_day_3","S_sdX_sdY_per_Pix_day_4","S_sdX_sdY_per_Pix_day_5",
             "S_sdX_sdY_per_Pix_AVG","pH46_sdX_sdY_per_Pix_day_1","pH46_sdX_sdY_per_Pix_day_2","pH46_sdX_sdY_per_Pix_day_3",
             "pH46_sdX_sdY_per_Pix_day_4","pH46_sdX_sdY_per_Pix_day_5","pH46_sdX_sdY_per_Pix_AVG",
             "Fe_GR_TL_day_1.2","Fe_GR_TL_day_2.3","Fe_GR_TL_day_3.4","Fe_GR_TL_day_4.5",
             "Fe_GR_TL_AVG","Fe_sdX_sdY_per_Pix_day_1","Fe_sdX_sdY_per_Pix_day_2","Fe_sdX_sdY_per_Pix_day_3",
             "Fe_sdX_sdY_per_Pix_day_4","Fe_sdX_sdY_per_Pix_day_5","Fe_sdX_sdY_per_Pix_AVG")
```

For each nutrient dataset, create a new table with the necessary columns log transformed.
```{r transform, eval=FALSE}
#Fe traits
t_fe_traits=data.frame(matrix(ncol = 93, nrow=2161))
names(t_fe_traits)=names(fe_traits)
for(i in 1:93){
  if(names(fe_traits)[i] %in% log_traits){
    t_fe_traits[,i]=log(fe_traits[,i])
    names(t_fe_traits)[i]=paste0("log_",names(fe_traits)[i])
  }
  else{t_fe_traits[,i]=fe_traits[,i]}
}

#USMS traits
t_usms_traits=data.frame(matrix(ncol = 94, nrow=5424))
names(t_usms_traits)=names(usms_traits)
for(i in 1:94){
  if(names(usms_traits)[i] %in% log_traits){
    t_usms_traits[,i]=log(usms_traits[,i])
    names(t_usms_traits)[i]=paste0("log_",names(usms_traits)[i])
  }
  else{t_usms_traits[,i]=usms_traits[,i]}
}

#Ph traits
t_ph_traits=data.frame(matrix(ncol = 90, nrow=2088))
names(t_ph_traits)=names(ph_traits)
for(i in 1:90){
  if(names(ph_traits)[i] %in% log_traits){
    t_ph_traits[,i]=log(ph_traits[,i])
    names(t_ph_traits)[i]=paste0("log_",names(ph_traits)[i])
  }
  else{t_ph_traits[,i]=ph_traits[,i]}
}

#P traits
t_p_traits=data.frame(matrix(ncol = 95, nrow=1710))
names(t_p_traits)=names(p_traits)
for(i in 1:95){
  if(names(p_traits)[i] %in% log_traits){
    t_p_traits[,i]=log(p_traits[,i])
    names(t_p_traits)[i]=paste0("log_",names(p_traits)[i])
  }
  else{t_p_traits[,i]=p_traits[,i]}
}

#S traits
t_s_traits=data.frame(matrix(ncol = 90, nrow=1558))
names(t_s_traits)=names(s_traits)
for(i in 1:90){
  if(names(s_traits)[i] %in% log_traits){
    t_s_traits[,i]=log(s_traits[,i])
    names(t_s_traits)[i]=paste0("log_",names(s_traits)[i])
  }
  else{t_s_traits[,i]=s_traits[,i]}
}

```

##Put transformed dataset into single matrix

Make a 'key' data table associating plant_id with ACC_ID and Treatment
```{r key, eval=FALSE}
key = t_p_traits$plant_id
key = rbind(key,t_usms_traits$plant_id)
key = rbind(key,t_fe_traits$plant_id)
key = rbind(key,t_s_traits$plant_id)
key = rbind(key,t_ph_traits$plant_id)
key=as.data.frame(key)
names(key)="plant_id"
key = separate(key,plant_id,sep="_",into=c("Treatment","ACC_ID","Rep"),remove=FALSE)
```

Move the plant_id column so it is the first in each data table
```{r first,eval=FALSE}
t_usms_traits = t_usms_traits[,-1]
t_usms_traits <- t_usms_traits %>%
  select(plant_id, everything())

t_p_traits = t_p_traits[,-1]
t_p_traits <- t_p_traits %>%
  select(plant_id, everything())

t_s_traits = t_s_traits[,-1]
t_s_traits <- t_s_traits %>%
  select(plant_id, everything())

t_ph_traits = t_ph_traits[,-1]
t_ph_traits <- t_ph_traits %>%
  select(plant_id, everything())

t_fe_traits = t_fe_traits[5:93]
t_fe_traits <- t_fe_traits %>%
  select(plant_id, everything())
```

Create an empty 12941x449 matrix and add in each data table in the order:
usms - p - s - ph - fe

```{r matrix, eval=FALSE}
#Make empty dataframe
t_all_traits = data.frame(matrix(ncol = 449, nrow=12941))
#Make a list of all the trait names from each dataframe
trait_names = names(t_usms_traits)[2:93]
trait_names = append(trait_names,names(t_p_traits)[2:94])
trait_names = append(trait_names,names(t_s_traits)[2:89])
trait_names = append(trait_names,names(t_ph_traits)[2:89])
trait_names = append(trait_names,names(t_fe_traits)[2:89])

#Create the plant_id column
plant_id = t_usms_traits$plant_id
plant_id = append(plant_id,t_p_traits$plant_id)
plant_id = append(plant_id,t_s_traits$plant_id)
plant_id = append(plant_id,t_ph_traits$plant_id)
plant_id = append(plant_id,t_fe_traits$plant_id)

#Add the trait names to the matrix columns
names(t_all_traits)=trait_names
t_all_traits$plant_id = plant_id
t_all_traits <- t_all_traits %>%
  select(plant_id, everything())

#Place in the matrix
t_all_traits[1:5424,2:93] = t_usms_traits[,2:93]
t_all_traits[5425:7134,94:186] = t_p_traits[,2:94]
t_all_traits[7135:8692,187:274] = t_s_traits[,2:89]
t_all_traits[8693:10780,275:362] = t_ph_traits[,2:89]
t_all_traits[10781:12941,363:450] = t_fe_traits[,2:89]
```


Some of the values in the dataframe are 'Inf' because it was divided by zero.
Change these values to NA.
```{r inf, eval=FALSE}
cf_DFinf2NA <- function(x)
{
  for (i in 1:ncol(x)){
    x[,i][is.infinite(x[,i])] = NA
  }
  return(x)
}

t_all_traits=cf_DFinf2NA(t_all_traits)
```

Finally, we can write these tables out to files, to run BSFG
```{r write, eval=FALSE}
write.table(t_all_traits,file='T_RootTraitMatrix.txt',sep='\t')
key$ACC_ID=as.factor(key$ACC_ID)
write.table(key,file='T_RootTraitPlantKey.txt',sep='\t')
```





